### 2019-09-29 boek.tex
update

% style packages
\usepackage{lmodern}
\usepackage{fancyhdr}
\usepackage{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion

@@

@@

%% hyperlinks
% load hyperref after fncychap
\usepackage{hyperref}
\usepackage[open,openlevel=1]{bookmark}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%        CH2: DYNGEN        %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dyngen: simulating single cells} 
\chaptermark{dyngen: simulating single cells}
\label{chap:dyngen}
\graphicspath{{./ch2_dyngen/}}
\include{ch2_dyngen/content}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%         CH4: DYNO         %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\newpage{\thispagestyle{empty}\cleardoublepage}
%\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
%\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
%\label{chap:dyno}
%\begin{quote}
%	\textbf{Abstract:} 
%\end{quote}
%\vfill
%Adapted from:\\
%\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.
%\include{ch4_dyno/content}

>>>>

% style packages
\usepackage{lmodern}
\usepackage{fancyhdr}
\usepackage{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
\usepackage{parskip} % disable paragraph indent and set paragraph skipping
\setlength{\emergencystretch}{3em}  % prevent overfull lines

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%&%%%
%%%      RMD packages      %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% use upquote for straight quotes in verbatim environments
\usepackage{upquote}

% fancy verbatim
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}

% common commands
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% something with figures
\usepackage{grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}

% something with making smaller itemizes
\providecommand{\tightlist}{%
	\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% booktab tables
\usepackage{longtable}
\usepackage{array}
\usepackage{wrapfig}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

@@

%% hyperlinks
% load hyperref after fncychap
\usepackage{hyperref}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[open,openlevel=1]{bookmark}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%        CH2: DYNGEN        %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dyngen: benchmarking with \textit{in silico} single cells} 
\chaptermark{dyngen: benchmarking with \textit{in silico} single cells}
\label{chap:dyngen}
\graphicspath{{./ch2_dyngen/}}
\include{ch2_dyngen/content}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%         CH3B: DYNO         %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
\label{chap:dyno}
\begin{quote}
	\textbf{Abstract:} Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory. Since 2014, at least 75 tools for trajectory inference have been developed. However, for end-users with a dataset of interest, these methods are difficult to execute and compare, mainly due to high variability in input/output data structures, software requirements, and programming interfaces.

	We developed dyno, a set of software tools that allow a user to easily infer, visualise and interpret single-cell trajectories. First, the user can select the most optimal set of methods based on the size of the dataset, the prior knowledge on the trajectories' topology and other user preferences. Each selected method can then be easily run within a common interface. Because the output of these methods are all converted into a common consistent format, the trajectories can be easily interpreted. For this, we provide a complete toolkit: adding a root and labelling the different cell stages, detecting genes which are differentially expressed at different stages of the trajectory, and plotting the trajectory within any dimensionality reduction or heatmap.

	We are currently extending this pipeline to include additional input data, such as RNA velocity measurements, and additional tools for interpreting the trajectories, such as alignment and differential expression.
\end{quote}
\vfill
Adapted from:\\
\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.\\
{\footnotesize $^*$ Equal contribution}
\newpage
\graphicspath{{./ch3b_dyno/}}
\include{ch3b_dyno/content}

### 2019-09-29 ch1_introduction/content.tex
update

Recent advances in experimental technologies are playing a crucial role in reductionist biology, allowing to measure the abundance of thousands of different biochemical molecules in tens of thousands of individual cells. With it comes the challenge of analysing what scientists refer to as “big data”, an amount of data that is not easily analysable and interpretable by hand. The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret. Namely, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}. Biologists thus turn to computer scientists
to develop new tools to tackle these problems and help them to extract meaningful biological insights from the data. In this work, incremental contributions were made to the field in order to be able to address the aforementioned problems in a more comprehensive context.

@@

%Technological advancements in single-cell omics allow studying a dynamic process in a high-throughput manner. 
%This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them \cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}. 
%Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place \cite{cannoodt_computationalmethodstrajectory_2016}. 
%TI has two objectives: to reconstruct the topology of the dynamic process (e.g. is it linear, cyclical, bifurcating), and to determine the position of each cell along the topology. Some TI methods assumes that the user knows the topology beforehand and only focuses on ordering the cells along a predefined topology.

@@

In this work, we aim to speed up scientific progress in single-cell omics by providing end-users with guidelines on how to use which state-of-the-art tools, and by providing developers with the necessary tools to assert a mimimum performance criterion when introducing new methods.

We develop benchmarking strategies for assessing the performance of novel computational tools constrained by low availability of real data (Chapter~\ref{chap:dyngen}). \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.

We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking.

We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}).
Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.

We conclude by inventing a new type of approach aimed at inferring the GRN of individual cells (Chapter~\ref{chap:bred}).
Due a lack of real data which can be used as a gold standard dataset, we use \textit{in silico} single-cell data to quantitatively evaluate the performance.

\mycomment{Add summaries for IncGraph, discussion, self-assessment and essential guidelines}

@@

\begin{itemize}
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
  \item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
  \item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
  \item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
  \item \textbf{Cannoodt R}, Saelens W, Saeys Y. bred: Inferring single cell regulatory networks. In preparation.
  \item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
 \end{itemize}

>>>>

Recent advances in experimental technologies are playing a crucial role in reductionist biology, allowing to measure the abundance of thousands of different biochemical molecules in tens of thousands of individual cells. With it comes the challenge of analysing a large amounts of data that are not easily interpretable by hand. The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret. For instance, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}. Biologists thus turn to computer scientists
to develop new tools to tackle these problems and help them to extract meaningful biological insights from the data. In this work, incremental contributions were made to the field in order to be able to address the aforementioned problems in a more comprehensive context.

@@

Technological advancements in single-cell omics allow studying a dynamic process in a high-throughput manner. 
This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them \cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}. 
Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place \cite{cannoodt_computationalmethodstrajectory_2016}. 
TI has two objectives: to reconstruct the topology of the dynamic process (e.g. is it linear, cyclical, bifurcating), and to determine the position of each cell along the topology. Some TI methods assumes that the user knows the topology beforehand and only focuses on ordering the cells along a predefined topology.

@@

In this work, we aim to speed up scientific progress in single-cell omics by providing tools both for end-users and developers alike. For developers of computational approaches, we provide tools and guidelines for benchmarking their method on real and synthetic data. For end-users we develop new tools and guidelines for analysing dynamic processes by inferring trajectories and gene regulatory networks. These contributions are discussed in the following chapters:

\begin{itemize}
  \item We develop benchmarking strategies for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}). \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
  \item We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking. We provide a set of guidelines for end-users wishing to infer trajectories. We also make our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
  \item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}). Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
  \item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
  \item Every NI method has certain topological biases. We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
  \item We discuss reproducibility problems of TI methods due to low rates of quantitative self-assessment (Chapter~\ref{chap:selfassessment}). We provide solutions for different causal reasons for this phenomenon in order to spur developers to perform more self-assessments.
  \item Finally, we summarise our experience in benchmarking computational methods in a list of essential guidelines (Chapter~\ref{chap:guidelines}).
\end{itemize}

@@

\begin{itemize}
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
  \item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
  \item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
  \item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
  \item \textbf{Cannoodt R}, Saelens W, Saeys Y. bred: Inferring single cell regulatory networks. In preparation.
  \item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
 \end{itemize}
